A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-17 tex/body.tex
more

Codatatypes are freely generated by their constructors, but in contrast with datatypes,
infinit\-e constructor terms are also legitimate values for codatatypes
(Section~\ref{sec:the-theory-of-co-datatypes}). Intuitively, the
values of a codatatype consist of all well-typed finite and infinite ground
constructor
terms, and only those. For example, the coinductive specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that
models the natural numbers $\const{Z}$, $\const{S}(\const{Z})$, $\const{S}(\const{S}(\const{Z}))$, $\ldots$\afterLdots{},
using Peano notation but extended with an
infinite value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$.
The equation $\const{S}(\infty) \teq \infty$ holds as expected,
because both sides expand to the infinite term
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies the
value~$\infty$.
Compared
with the conventional inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Infty}\]
with a dedicated constructor for representing infinity, the codatatype avoids
one case by unifying the finite and infinite nonzero cases.

@@

\pagebreak[3]

@@

\begin{rep}\paragraph{Related Work.} \end{rep}%
Although it was written a few years ago,
Barrett et al.\ \cite{barrett-et-al-2007} still provides a good account of
related work. Since then, datatypes have been added to the SMT solver
Z3 in unpublished work \begin{rep}by Leonardo de Moura \end{rep}and to a SPASS-like %prototypical
superposition prover \begin{rep}by Daniel Wand\end{rep} \cite{wand-2014}.
Closely related is the automatic structural
induction in superposition provers \cite{kersani-peltier-2013,wand-weidenbach-201x}
and SMT solvers \cite{reynolds-kuncak-2015}.
%
Datatypes are explicitly supported by the higher-order model finder Refute
\cite{weber-2008}. Its successor, Nitpick \cite{blanchette-nipkow-2010}, can
also find models involving cyclic codatatype values. The $\mu$-notation is
inspired by the $\mu$-calculus \cite[etc.]{kozen-1983,endrullis-et-al-2011}.

@@

%\paragraph{Specifications.}
In an SMT problem, the signature is normally given by specifying first the
uninterpreted types in any order, then the (co)datatypes with their constructors
and selectors in groups of mutually (co)recursive groups of (co)datatypes, and
finally any other function symbols.
%
Each (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures fall outside this fragment%
\dotReportFootnote{%
In principle, rank-1 (top-level) polymorphism \cite{blanchette-paskevich-2013}
should not raise any special difficulties. Nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).}
In the presentation, we allow ourselves some metalevel parameterization
% at the metalevel
through subscripts---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types including
$\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
and \smash{$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$}.

>>>>

Codatatypes are freely generated by their constructors, but in contrast with datatypes,
infinit\-e constructor terms are also legitimate values for codatatypes
(Section~\ref{sec:the-theory-of-co-datatypes}). Intuitively, the
values of a codatatype consist of all well-typed finite and infinite ground
constructor
terms, and only those. For example, the coinductive specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that
models the natural numbers $\const{Z}$, $\const{S}(\const{Z})$, $\const{S}(\const{S}(\const{Z}))$, $\ldots$\afterLdots{},
using Peano notation but extended with an
infinite value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$.
The equation $\const{S}(\infty) \teq \infty$ holds as expected,
because both sides expand to the infinite term
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies the
value~$\infty$.
Compared
with the conventional
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Infty}\]
with a dedicated constructor for representing infinity, the codatatype avoids
one case by unifying the finite and infinite nonzero cases.

@@

@@

\begin{rep}\paragraph{Related Work.} Although\end{rep}%
\begin{conf}Although\end{conf} it was written a few years ago,
Barrett et al.\ \cite{barrett-et-al-2007} still provides a good account of
related work. Since then, datatypes have been added to the SMT solver
Z3 in unpublished work \begin{rep}by Leonardo de Moura \end{rep}and to a SPASS-like %prototypical
superposition prover \begin{rep}by Daniel Wand\end{rep} \cite{wand-2014}.
Closely related is the automatic structural
induction in superposition provers \cite{kersani-peltier-2013,wand-weidenbach-201x}
and SMT solvers \cite{reynolds-kuncak-2015}.
%
Datatypes are explicitly supported by the higher-order model finder Refute
\cite{weber-2008}. Its successor, Nitpick \cite{blanchette-nipkow-2010}, can
also find models involving cyclic codatatype values. The $\mu$-notation is
inspired by the $\mu$-calculus \cite[etc.]{kozen-1983,endrullis-et-al-2011}.

@@

%\paragraph{Specifications.}
In an SMT problem, the signature is normally given by specifying first the
uninterpreted types in any order, then the (co)datatypes with their constructors
and selectors in groups of mutually (co)recursive groups of (co)datatypes, and
finally any other function symbols.
%
Each (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures fall outside this fragment%
\dotReportFootnote{%
In principle, rank-1 (top-level) polymorphism \cite{blanchette-paskevich-2013}
should not raise any special difficulties. Nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).}
In the presentation, we allow ourselves some metalevel parameterization
% at the metalevel
through subscripts%
\begin{conf}
(e.g., $\ty{stream}_{\,\tau}$)%
\end{conf}%
\begin{rep}---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types including
$\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
and \smash{$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$}\end{rep}.
